Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-27 doc/papers/cgo2014/paper.tex
Some info on expression templates

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, float, label={lst:l2norm-loopy}]
def slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, float, label={lst:l2norm}]
import numpy as np
def l2norm(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

@@

\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, float, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

@@

\subsection{Temporaries Elimination}

@@

\subsection{Operator Fusion}

@@

%%
%%
\section{Compiler - Runtime Cooperation}
\label{sec:pythran}

\subsection{Typing}

@@

\cite{expression_templates, et2012}

>>>>

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, label={lst:l2norm-loopy}]
def slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, label={lst:l2norm}]
import numpy as np
def l2norm(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

@@

\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

@@

\subsection{Temporaries Elimination}
\label{sec:temporaries-elimination}

@@

\subsection{Operator Fusion}
\label{sec:operator-fusion}

@@

%%
%%
\section{Compilers Cooperation}
\label{sec:pythran}

C/C++ have been at the hearth of high performance computing for many years.
Compiler infrasturctures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{boostsimd} provide good
abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.

There is no need to reinvent the wheel when compiling numerical Python kernels.
A more profitable approach is to map Python concept to existing ones, or
translate them into existing them. As a consequence, this section focuses on
the translation of Python constructs to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.

@@

Many C++ linear algebra libraries~\cite{eigen,ublas} use \emph{expression
templates}~\cite{expression_templates, et2012} as a way to efficently describe high
level matrix expressions while avoid the creation of intermediate storage.
Additionnaly, it has the same effect as loop fusion, solving the issues
described in \S~\cite{sec:temporaries-elimination} and
\S~\ref{sec:operator-fusion}.

Thanks to variadic templates introduced in C++11, it is possible to write a
generic C++ meta-class that implements the same API as numpy array and provides
all unary and binary operators, as well as the various form of indexing and
slicing. Listing~\ref{lst:rosen-cxx} illustrates the translation of the Python
code from Listing~\ref{lst:rosen} into C++. Altough this version is more
verbose than the Python one, the translation from one to the other does not
raise any particular issue.

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
constexpr size_t N = x::ndim;
ndarray<N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}
